NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-09-01 paper/Chapters/Chapter3.tex
grammarly checked

A series of traditional LCA of the responses to carbohydrate intake within the 7 time slots of the day was first examined. These initial analyses ignored the clustering of observation days within participants of the survey. \textbf{Table \ref{tab:mixmodels}} shows the latent class solutions for one to five classes (see rows under the Fixed effects model section). The BIC declines as the number of day level classes increases. However, the improvement of BIC dropped to less than 1000 from 3 class to 4 class solutions (658.9) and from 4 class to 5 class solutions (361.7). Entropy index indicates that the 4 class model could explain about 51\% percent of the data, while \textit{p} values of Lo-Mendell-Rubun LRT suggest that the more classes we fit, the better model we will have until up to 6 classes (\textit{p} = 0.06 and is not shown in the table). For the sake of parsimony, we only extended the model with random effects building on 2 class, 3 class, and 4 class solutions.

The results of the random effects models are presented in \textbf{Table \ref{tab:mixmodels}} under the Random effects model section. It is obvious that the BIC improves with the addition of random effects to account for the nested structure of the data. Entropy indicates that 2 classes in the day level and 4 classes in the individual level may be the best solution mathematically. However, after these solutions were checked in more details, the most substantively interpretable model was found to be the 3$\times$3 random effect model, which is the model with 3 latent classes in the day level and 3 latent classes in the individual level. We must emphasize that different researchers may have made decisions slightly different from ours, we have provided other solutions in \textbf{Appendix \ref{AppendixC}} for reference.

@@

As expected, consumption of total energy (7539.98 kJ), total carbohydrate (222.79 g), total sugar (98.63 g), starch (124.07 g), and non-milk extrinsic sugar (59.45 g) were the highest among high percentage carbohydrate days (all \textit{p} < 0.001). On the other hand, the consumption of protein (73.29 g), total fat (67.24 g), and alcohol (13.80 g) were the highest in the regular meals days. Moreover, in the high percentage carbohydrate days, participants turned out to consume the highest amount of fruit (107.40 g). There was no evidence of any difference in the consumption of yellow, red, or green vegetables across the three types of days (\textit{p} = 0.681).\vspace{-0.6cm}

@@

In the random effect models, we utilized the non-parametric approach, in which we added a level 2 (individual level) latent classes based on the random means from the level 1 (day level) latent class solution. The results of the individual level LCA solution for 2 and 3 classes are presented in \textbf{Figure \ref{fig:CB2level2}}, and \textbf{\ref{fig:level2}}.

@@

With three individual-level latent classes \textbf{(Figure \ref{fig:level2})}, a low-carbohydrate eaters class, a moderate-carbohydrate eaters class, and a high-carbohydrate eaters class emerge. 43.1\% participants were identified as high-carbohydrate eaters, in these individuals, about 50\% of the days (2 out of 4 days) of their dietary diary could be classified as having high percentage carbohydrate days. Nearly 1 out of 4 days of their dietary diary were either "regular meals day" or "low carbohydrate day". 28.1\% participants fell into the low carbohydrate eaters class on the left-hand side of \textbf{Figure \ref{fig:level2}}, their recordings of food intake showed that in more than 60\% of their days, they were having "regular meals" which was characterised as with the highest amount of fat and alcohol consumptions as already described in \textbf{Table \ref{tab:day-level-features}}. Moderate carbohydrate eaters have comparable proportions (42.0\% vs. 40.0\%) of high carbohydrate days and regular meals days; 18.0\% of their dietary diary was found to be low carbohydrate days.

@@

Among the three types of carbohydrate eaters, the mean of total energy intake over the 4 days of dietary survey was the highest (7985.8 kJ, 95\%CI: 7283.3, 8146.3) in the low carbohydrate eaters group, and the lowest (7341.8 kJ, 95\%CI: 7172.5, 7511.2) in the moderate eaters group \textbf{(Table \ref{tab:tab-nutri-indi})}. Sources of energy for each type of carbohydrate eaters by the 7-time slots were also different. Low carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-A)} never had carbohydrate contributing more than 50\% of their total energy throughout the day. Energy from fat was the highest for low carbohydrate eaters most of the time during the day (except for the time slot between 10 pm to 6 am next morning). Most impressively, energy contributed by alcohol was always the highest in low carbohydrate eaters, percentages for energy from alcohol for the 7-times slots were 1.4\% (6-9 am), 1.8\% (9-12 noon), 3.1\% (12-2 pm), 7.2\% (2-5 pm), 8.4\% (5-8 pm), 20.0\% (8-10 pm), and 31.2\% (10-6 am), respectively. Contributions from different energy sources were quite similar for moderate and high carbohydrate eaters, but their absolute amount of energy consumption at each time slot was largely different. Moderate carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-B)} were characterised as consuming the lowest energy (296.9 kJ) before 9 am but having higher energy consumption (967.9 kJ) between 9 am and 12 noon time compared with low and high carbohydrate eaters. Moderate carbohydrate eaters may tend to have later breakfast, later lunch, and probably later dinner as well. They had the highest total energy consumption (599.7 kJ) at night (10 pm - 6 am) across all three types of eaters. High carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-C)} consumed the highest total energy (929.0 kJ) during 6 am to 9 am in the morning and the lowest total energy between 10 pm to 6 am (205.5 kJ). Specifically, carbohydrate contributions to total energy intake were 65.3\% (6-9 am), 56.2\% (9-12 noon), 46.2\% (12-2 pm), 50.3\% (2-5 pm), 42.5\% (5-8 pm), 44.1\% (8-10 pm), and 49.9\% (10-6 am). We also noticed that high carbohydrate eaters consumed their energy mainly from the following three time slots: 6-9 am, 12-2 pm, and 5-8 pm.

In total, the mean of total carbohydrate intake was 203.8 g, 218.3 g, and 233.4 g for low, moderate, and high carbohydrate eaters, respectively \textbf{(Table \ref{tab:tab-nutri-indi})}. Total energy contributed by carbohydrate was close to 50\% among the high carbohydrate eaters but was only 40.6\% among the low carbohydrate eaters. In terms of the subtypes (components) of the carbohydrate consumed at each time slot, high carbohydrate eaters consumed more than 2 times (compared to low carbohydrate eaters) and nearly 4 times (against moderate carbohydrate eaters) the amount of sugar (37.9g 95\% CI: 36.8, 39.2) and non-milk extrinsic sugar (i.e. free sugar, 11.1g 95\%CI: 10.7, 11.6) between 6-9 am. Moderate carbohydrate eaters had carbohydrate intakes that were more spread out throughout the day. They consumed more sugar and starch during 9-12 noon, 2-5 pm, 8-10 pm, and 10-6 am. Low carbohydrate eaters turned to have similar temporal pattern of consuming carbohydrates as the high carbohydrate eaters, but the absolute amount of fibre, sugar, free sugar, and starch consumed were usually lower than that in the high carbohydrate eaters except for time slots of 2-5 pm, and 10-6 am. Strong evidence (\textit{p} < 0.001) suggested that the

@@

The social-demographic characteristics of the UK adults according to their individual level latent class membership are shown in \textbf{Table \ref{tab:Level2tab1}}. Moderate carbohydrate eaters were relatively younger (\textit{p} < 0.001). Gender distribution across the three types of carbohydrate eaters was fairly even (\textit{p} = 0.119). The distribution of the carbohydrate eater types appears to be changing with the year of survey. Low carbohydrate eaters represented 32.5\% of the population in the first year of the survey but later dropped to lower than 30\% (minimum in the third year at 22.6\%) until the most recent year. The proportion of high carbohydrate eaters increased from 41.2\% to the highest (50.6\%) in the second year, but then started to decline to 38.4\% in the 8th year of survey (\textit{p} = 0.015). There was no evidence of any difference in employment status across three types of carbohydrate eaters. However, strong evidence suggested that high carbohydrate eaters had the highest proportion (61.3\%) of living with partner (\textit{p} < 0.001); moderate carbohydrate eaters had the lowest average income (27180.8 \textsterling/year), the highest proportion of non-white population (20.5\%), and a lower education level (23.3\% with degree of higher education) compared with either low or high carbohydrate eaters.

@@

\textbf{Table \ref{tab:tab1hypetension}} presents the characteristics of men and women participants in the NDNS RP 2008/09-15/16 by hypertension status. The weighted prevalences of hypertension were 30.4\% in men and 27.5\% in women. Among both sexes, there was strong evidence of differences by hypertension status for age, education level, living with a partner or not, smoking status, BMI, abdominal obesity (WC), and prevalence of diabetes (\textit{p} < 0.01). No difference was found among either men or women for ethnicity. Strong evidence of difference was suggested in women for average household income (32741.5 \textsterling/year in non-hypertensive vs. 27862.0 \textsterling/year in hypertensive, \textit{p} < 0.001), and physical activity level (geometric mean: 0.81 hours/day in non-hypertensive compared with 0.53 hours/day in hypertensive, \textit{p} < 0.001) but not in men. Interestingly, in both sexes, hypertensive participants had a higher proportion of being classified as low or high carbohydrate eaters; total energy and carbohydrate intake were higher in people without hypertension (\textit{p} < 0.001).

@@

\textbf{Table \ref{tab:tab1BMI}} shows the characteristics of participants according to their obesity status stratified by sex. The survey design-weighted prevalence for being overweight and obese in the UK adults were estimated to be 43.4\% and 25.7\% in men, and 30.9\% and 27.4\% in women. Obviously, abdominal obesity (WC) increased significantly with the elevated BMI level in both men and women. Overweight or obese participants were older, having lower total energy intake and lower carbohydrate intake compared with normal weight men and women (\textit{p} < 0.001). Moreover, education

@@

Similarly, when looking at the association between carbohydrate eating pattern and abdominal obesity (WC), men who were classified as moderate carbohydrate eaters were found to have about 3 cm (95\%CI: 0.52, 5.49 cm, \textit{p} = 0.018) smaller WC compared to low carbohydrate eaters in the crude model \textbf{(Table \ref{tab:tab2WC})}. However, after adjustment for age, living with partner or not, average household income, education level, hypertension, smoking status, total energy intake, and alcohol consumption, the association attenuated to no difference. The interaction effect of living with a partner or not on the association between carbohydrate eating patterns and WC was again found in women in total (\textit{p} for interaction = 0.009) and without diabetes (\textit{p} for interaction = 0.012). Among women who were living with their partners, high carbohydrate eaters had 4.71 cm (95\%CI: 2.43, 7.00, \textit{p} < 0.001) smaller WC on average compared with low carbohydrate eaters. The association remained after restricting the sample to non-diabetic women (-3.74 cm, 95\% CI: -5.97, -1.51,  \textit{p} = 0.001). However, for women who were living alone, moderate carbohydrate eaters had 3.17 cm (95\%CI: 0.05, 6.30, \textit{p} = 0.047) larger WC on average compared with low carbohydrate eaters. The evidence for the positive association between moderate carbohydrate eaters and WC in women became weaker, but still had borderline significance when excluding diabetic women (3.08 cm, 95\%CI: -0.09, 6.25, \textit{p} = 0.057).

>>>>

A series of traditional LCA of the responses to carbohydrate intake within the 7 time slots of the day was first examined. These initial analyses ignored the clustering of observation days within participants of the survey. \textbf{Table \ref{tab:mixmodels}} shows the latent class solutions for one to five classes (see rows under the Fixed effects model section). The BIC declines as the number of day level classes increases. However, the improvement of BIC dropped to less than 1000 from 3 class to 4 class solutions (658.9) and from 4 class to 5 class solutions (361.7). Entropy index indicates that the 4 class model could explain about 51\% of the data, while \textit{p} values of Lo-Mendell-Rubun LRT suggest that the more classes we fit, the better model we will have until up to 6 classes (\textit{p} = 0.06 and is not shown in the table). For the sake of parsimony, we only extended the model with random effects building on 2 class, 3 class, and 4 class solutions.

The results of the random effects models are presented in \textbf{Table \ref{tab:mixmodels}} under the Random effects model section. It is evident that the BIC improves with the addition of random effects to account for the nested structure of the data. Entropy indicates that 2 classes in the day level and 4 classes in the individual level may be the best solution mathematically. However, after these solutions were checked in more details, the most substantively interpretable model was found to be the 3$\times$3 random effect model, which is the model with 3 latent classes in the day level and 3 latent classes in the individual level. We must emphasise that different researchers may have made decisions slightly different from ours, we have provided other solutions in \textbf{Appendix \ref{AppendixC}} for reference.

@@

As expected, consumption of total energy (7539.98 kJ), total carbohydrate (222.79 g), total sugar (98.63 g), starch (124.07 g), and non-milk extrinsic sugar (59.45 g) was the highest among high percentage carbohydrate days (all \textit{p} < 0.001). On the other hand, the consumption of protein (73.29 g), total fat (67.24 g), and alcohol (13.80 g) were the highest in the regular meals days. Moreover, in the high percentage carbohydrate days, participants turned out to consume the highest amount of fruit (107.40 g). There was no evidence of any difference in the consumption of yellow, red, or green vegetables across the three types of days (\textit{p} = 0.681).\vspace{-0.6cm}

@@

In the random effect models, we utilised the non-parametric approach, in which we added a level 2 (individual level) latent classes based on the random means from the level 1 (day level) latent class solution. The results of the individual level LCA solution for 2 and 3 classes are presented in \textbf{Figure \ref{fig:CB2level2}}, and \textbf{\ref{fig:level2}}.

@@

With three individual-level latent classes \textbf{(Figure \ref{fig:level2})}, a low-carbohydrate eaters class, a moderate-carbohydrate eaters class, and a high-carbohydrate eaters class emerge. 43.1\% participants were identified as high-carbohydrate eaters, in these individuals, about 50\% of the days (2 out of 4 days) of their dietary diary could be classified as having high percentage carbohydrate days. Nearly 1 out of 4 days of their dietary diary were either "regular meals day" or "low carbohydrate day". 28.1\% participants fell into the low carbohydrate eaters class on the left-hand side of \textbf{Figure \ref{fig:level2}}, their recordings of food intake showed that in more than 60\% of their days, they were having "regular meals" which was characterised as with the highest amount of fat and alcohol consumptions as already described in \textbf{Table \ref{tab:day-level-features}}. Moderate carbohydrate eaters have comparable proportions (42.0\% vs 40.0\%) of high carbohydrate days and regular meals days; 18.0\% of their dietary diary was found to be low carbohydrate days.

@@

Among the three types of carbohydrate eaters, the mean of total energy intake over the 4 days of dietary survey was the highest (7985.8 kJ, 95\%CI: 7283.3, 8146.3) in the low carbohydrate eaters group, and the lowest (7341.8 kJ, 95\%CI: 7172.5, 7511.2) in the moderate eaters group \textbf{(Table \ref{tab:tab-nutri-indi})}. Sources of energy for each type of carbohydrate eaters by the 7-time slots were also different. Low carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-A)} never had carbohydrate contributing more than 50\% of their total energy throughout the day. Energy from fat was the highest for low carbohydrate eaters most of the time during the day (except for the time slot between 10 pm to 6 am next morning). Most impressively, energy contributed by alcohol was always the highest in low carbohydrate eaters, percentages for energy from alcohol for the 7-times slots were 1.4\% (6-9 am), 1.8\% (9-12 noon), 3.1\% (12-2 pm), 7.2\% (2-5 pm), 8.4\% (5-8 pm), 20.0\% (8-10 pm), and 31.2\% (10-6 am), respectively. Contributions from different energy sources were quite similar for moderate and high carbohydrate eaters, but their absolute amount of energy consumption at each time slot was largely different. Moderate carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-B)} were characterised as consuming the lowest energy (296.9 kJ) before 9 am but having higher energy consumption (967.9 kJ) between 9 am and 12 noon time compared with low and high carbohydrate eaters. Moderate carbohydrate eaters may tend to have later breakfast, later lunch, and probably later dinner as well. They had the highest total energy consumption (599.7 kJ) at night (10 pm - 6 am) across all three types of eaters. High carbohydrate eaters \textbf{(Figure \ref{fig:energysourcesCB}-C)} consumed the highest total energy (929.0 kJ) during 6 am to 9 am in the morning, and the lowest total energy between 10 pm to 6 am (205.5 kJ). Specifically, carbohydrate contributions to total energy intake were 65.3\% (6-9 am), 56.2\% (9-12 noon), 46.2\% (12-2 pm), 50.3\% (2-5 pm), 42.5\% (5-8 pm), 44.1\% (8-10 pm), and 49.9\% (10-6 am). We also noticed that high carbohydrate eaters consumed their energy mainly from the following three time slots: 6-9 am, 12-2 pm, and 5-8 pm.

In total, the mean of total carbohydrate intake was 203.8 g, 218.3 g, and 233.4 g for low, moderate, and high carbohydrate eaters, respectively \textbf{(Table \ref{tab:tab-nutri-indi})}. Total energy contributed by carbohydrate was close to 50\% among the high carbohydrate eaters but was only 40.6\% among the low carbohydrate eaters. In terms of the subtypes (components) of the carbohydrate consumed at each time slot, high carbohydrate eaters consumed more than 2 times (compared to low carbohydrate eaters) and nearly 4 times (against moderate carbohydrate eaters) the amount of sugar (37.9g 95\% CI: 36.8, 39.2) and non-milk extrinsic sugar (i.e. free sugar, 11.1g 95\%CI: 10.7, 11.6) between 6-9 am. Moderate carbohydrate eaters had carbohydrate intakes that were more spread out throughout the day. They consumed more sugar and starch during 9-12 noon, 2-5 pm, 8-10 pm, and 10-6 am. Low carbohydrate eaters turned to have a similar temporal pattern of consuming carbohydrates as the high carbohydrate eaters, but the absolute amount of fibre, sugar, free sugar, and starch consumed were usually lower than that in the high carbohydrate eaters except for time slots of 2-5 pm, and 10-6 am. Strong evidence (\textit{p} < 0.001) suggested that the

@@

The social-demographic characteristics of the UK adults according to their individual level latent class membership are shown in \textbf{Table \ref{tab:Level2tab1}}. Moderate carbohydrate eaters were relatively younger (\textit{p} < 0.001). Gender distribution across the three types of carbohydrate eaters was reasonably even (\textit{p} = 0.119). The distribution of the carbohydrate eater types appears to be changing with the year of survey. Low carbohydrate eaters represented 32.5\% of the population in the first year of the survey but later dropped to lower than 30\% (minimum in the third year at 22.6\%) until the most recent year. The proportion of high carbohydrate eaters increased from 41.2\% to the highest (50.6\%) in the second year, but then started to decline to 38.4\% in the 8th year of survey (\textit{p} = 0.015). There was no evidence of any difference in employment status across three types of carbohydrate eaters. However, strong evidence suggested that high carbohydrate eaters had the highest proportion (61.3\%) of living with partner (\textit{p} < 0.001); moderate carbohydrate eaters had the lowest average income (27180.8 \textsterling/year), the highest proportion of non-white population (20.5\%), and a lower education level (23.3\% with degree of higher education) compared with either low or high carbohydrate eaters.

@@

\textbf{Table \ref{tab:tab1hypetension}} presents the characteristics of men and women participants in the NDNS RP 2008/09-15/16 by hypertension status. The weighted prevalences of hypertension were 30.4\% in men and 27.5\% in women. Among both sexes, there was strong evidence of differences by hypertension status for age, education level, living with a partner or not, smoking status, BMI, abdominal obesity (WC), and prevalence of diabetes (\textit{p} < 0.01). No difference was found among either men or women for ethnicity. Strong evidence of difference was suggested in women for average household income (32741.5 \textsterling/year in non-hypertensive vs 27862.0 \textsterling/year in hypertensive, \textit{p} < 0.001), and physical activity level (geometric mean: 0.81 hours/day in non-hypertensive compared with 0.53 hours/day in hypertensive, \textit{p} < 0.001) but not in men. Interestingly, in both sexes, hypertensive participants had a higher proportion of being classified as low or high carbohydrate eaters; total energy and carbohydrate intake were higher in people without hypertension (\textit{p} < 0.001).

@@

\textbf{Table \ref{tab:tab1BMI}} shows the characteristics of participants according to their obesity status stratified by sex. The survey design-weighted prevalences for being overweight and obese in the UK adults were estimated to be 43.4\% and 25.7\% in men, and 30.9\% and 27.4\% in women. Obviously, abdominal obesity (WC) increased significantly with the elevated BMI level in both men and women. Overweight or obese participants were older, having lower total energy intake and lower carbohydrate intake compared with normal weight men and women (\textit{p} < 0.001). Moreover, education

@@

Similarly, when looking at the association between carbohydrate eating pattern and abdominal obesity (WC), men who were classified as moderate carbohydrate eaters were found to have about 3 cm (95\%CI: 0.52, 5.49 cm, \textit{p} = 0.018) smaller WC compared to low carbohydrate eaters in the crude model \textbf{(Table \ref{tab:tab2WC})}. However, after adjustment for age, living with partner or not, average household income, education level, hypertension, smoking status, total energy intake, and alcohol consumption, the association attenuated to no difference. The interaction effect of living with a partner or not on the association between carbohydrate eating patterns and WC was again found in women in total (\textit{p} for interaction = 0.009) and without diabetes (\textit{p} for interaction = 0.012). Among women who were living with their partners, high carbohydrate eaters had 4.71 cm (95\%CI: 2.43, 7.00, \textit{p} < 0.001) smaller WC on average compared with low carbohydrate eaters. The association remained after restricting the sample to non-diabetic women (-3.74 cm, 95\% CI: -5.97, -1.51,  \textit{p} = 0.001). However, for women who were living alone, moderate carbohydrate eaters had 3.17 cm (95\%CI: 0.05, 6.30, \textit{p} = 0.047) larger WC on average compared with low carbohydrate eaters. The evidence for the positive association between moderate carbohydrate eaters and WC in women became weaker but still had borderline significance when excluding diabetic women (3.08 cm, 95\%CI: -0.09, 6.25, \textit{p} = 0.057).
